Statistical Inference

### 2025-03-04 paper-vis-inference/paper.qmd
Changes before proofread

---
title: "Statistical Inference"
subtitle: "A comparative review of Frequentist, Bayesian, and Visual methods"
author: Tyler Wiederich
date: "`r format(Sys.time(), '%B %d, %Y')`"
format: 
  wordcount-pdf: default
bibliography: references.bib
biblio-style: https://www.zotero.org/styles/apa
execute: 
  echo: false
  message: false
  warning: false
---

@@

The quantification of uncertainty is one of the key components behind statistical methodology [@acree2021; @lehmann2008book]. This is useful in cases where population parameters are unknown or unable to be reasonably measured. For example, some possible research questions that would require statistical methods include “What percentage of Minnesotan residents agree with the proposed legislation?,” or “Does the placement of higher margin products on the middle shelf of a store aisle lead to more profits than placement on top and bottom shelves?” These types of questions can be approximated using statistical inference, although the approach differs by the specific method of inference.

Implicitly, many statistical methods revolve around the concept of a hypothesis test [@neyman1933; @acree2021]. These tests are designed to determine whether or not there is evidence to support a claim. Data is evaluated against a null hypothesis, which is typically a baseline assumption or a specific condition. If the condition is deemed unlikely to occur by chance, then there is evidence to support its complement, called the alternative hypothesis. The strength of evidence can be quantified, leading to decisions about what interpretation or decision to make regarding the results of the method.

@@

The use of visualizations is a valuable method for understanding the structure and patterns in data [@tukeyEDA], but they can also guide formal analyses [@loy2015; @loy2016; @loy2017]. Visual inference is where hypotheses can be tested by human perception instead of mathematical formulation. For example, Q-Q plots are a common diagnostic tool for checking normality assumptions [@loy2016]. While methods like the Shapiro-Wilk test [@shapiro1965] exist for testing normality, visual inference can be applied when assumptions are violated or tests do not exist, as long as a reference model is obtainable for null data generation.

@@

In the early development of Frequentist statistics, philosophical debates between R. A. Fisher and J. Neyman led to the distinction between "inductive inference" and "inductive behavior" [@lehmann2008]. Inductive inferences, as preferred by Fisher, involves logical rational to develop models. On the other hand, Neyman's inductive behavior addresses the additional component of randomization through hypothesis testing. Both arguments start with the assumption that the random variable in question is independent and identically distributed. However, the two statisticians differed on the interpretations and approaches, with Fisher favoring intuition and Neyman favoring proofs. The methodology was similar between them, emphasizing specific approaches and scope of inference.

@@

```{r}
#| label: fig-freq-eval-methods
#| fig-cap: "Frequentist evaluation methods using test statistics and confidence intervals under a null hypothesis that uses a standard normal distribution. Panel (a) shows the area under a standard normal for the Z-score of the observed sample mean. Panel (b) provides 95% confidence intervals for Z-scores of the sample mean across multiple samples."
#| fig-subcap: 
#| - "Probability of observing test statistic under null hypothesis."
#| - "95% confidence interval for multiple repeated samples."
#| out-width: 90%

set.seed(20244)
ggplot() + 
  stat_function(fun = dnorm) + 
  stat_function(fun = dnorm, xlim = c(1.5,4), 
                geom = 'area',
                fill = 'red', color = 'black') + 
  labs(x = 'Z', y = expression(phi(Z))) + 
  scale_x_continuous(limits = c(-3, 3)) + 
  theme(aspect.ratio = 1)

freq.ci <- data.frame()
for(i in 1:100){
  t.test.save <- t.test(rnorm(100))
  tmp <- data.frame(mean = t.test.save$estimate,
                    lcl = t.test.save$conf.int[1],
                    ucl = t.test.save$conf.int[2],
                    id = i)
  freq.ci <- bind_rows(freq.ci, tmp)
  rm(t.test.save)
  rm(tmp)
}
freq.ci %>% 
  mutate(contain0 = ifelse(0 > lcl & 0 <= ucl, 'yes', 'no')) %>% 
  ggplot(mapping = aes(x = mean, y = id, color = contain0)) + 
  geom_point() + 
  geom_errorbar(aes(xmin = lcl, xmax = ucl)) + 
  scale_color_brewer(palette = 'Set1') +
  labs(x = 'Z', y = 'Sample ID', color = 'CI contain zero?') + 
  geom_vline(xintercept = 0, color = 'red', linetype = 'dashed') + 
  theme(aspect.ratio = 1)
```

@@

```{r}
#| label: fig-bayes-binom
#| fig-cap: "Caption"
#| fig-subcap: 
#| - "a: likelihoods"
#| - "b: data simulated from models"
#| layout-ncol: 2
#| out-width: 90%

@@

tibble(
  x = seq(0,1, l = 1000)
) %>% 
  mutate(fx.u = dbeta(x, 13+1, 20-13+1),
         fx.b = dbeta(x, 13+alpha+1, 20-13+beta+1),
         L = choose(20,13)*x^(13)*(1-x)^(20-13)) %>% 
  ggplot(mapping = aes(x = x)) + 
  geom_line(aes(y = fx.u, linetype = 'Uniform')) + 
  geom_line(aes(y = fx.b, linetype = 'Beta')) +
  geom_vline(xintercept = 0.7, color = 'red') +
  scale_linetype_manual(values = prior) + 
  labs(x = 'p', y = 'Posterior density', linetype = 'Prior')

@@

wild <- read_csv('wild23-25.csv', show_col_types = F) %>% 
  mutate(pt_diff = abs(gf-ga))
wild %>% 
  group_by(season, result, location) %>% 
  count() %>% 
  pivot_wider(names_from = location, values_from = n) %>% 
  mutate(season = paste0(season-1, '-', season)) %>% 
  knitr::kable(caption = 'Data from the 2023-24 and 2024-25 seasons for the Minnesota Wild. 0 represents losses and 1 represents wins.')
```

@@

```{r}
#| label: fig-true-props
#| fig-height: 4
#| fig-width: 6
#| fig-cap: ""

wild %>% 
  group_by(season, location, result, pt_diff) %>% 
  count() %>% 
  pivot_wider(names_from = result, names_prefix = 'result',
              values_from = n, values_fill = 0) %>% 
  mutate(propwin = result1 / (result0+result1)) %>% 
  filter(!(result0 == 0 & result1 == 0)) %>% 
  ggplot(mapping = aes(x = pt_diff, y = propwin, color = location)) + 
  geom_line() + 
  geom_point() + 
  scale_y_continuous(limits = c(0,1)) + 
  labs(x = 'Point differential',
       y = 'Proportion of wins') +
  scale_color_manual(values = c('home' = '#154734', 'away' = '#A6192E')) +
  facet_wrap(~season) + 
  theme_bw() + 
  theme(aspect.ratio = 1) 
```

@@

Fitting a Frequentist generalized linear model (GLM) as shown in @eq-freq-glm reveals that the point differential is the only significant term (@tbl-anova), which means that the Wild do not seem to have a winning advantage by playing on home-ice. The only conclusion from the this analysis is that there is evidence that the Minnesota Wild are less likely to win games for larger point differentials than for smaller point differentials.

@@

```{r}
#| label: tbl-coef-compare
data.frame(
  Frequentist = coef(mod.freq),
  `Flat priors` = fixef(bayes.unif)[,1],
  `Informative priors` = fixef(bayes.priors)[,1]
) %>% 
  rownames_to_column(var = 'Term') %>% 
  knitr::kable(digits = 3, caption = 'Coefficients of Frequentist and Bayesian methods for GLMs fit to the Minnesota Wild data.')

>>>>

---
title: "Statistical Inference"
subtitle: "A comparative review of Frequentist, Bayesian, and Visual methods"
author: Tyler Wiederich
date: "`r format(Sys.time(), '%B %d, %Y')`"
format: wordcount-pdf
bibliography: references.bib
biblio-style: https://www.zotero.org/styles/apa
execute: 
  echo: false
  message: false
  warning: false
  include: false
---

@@

The quantification of uncertainty is one of the fundamental concepts behind statistical methodology [@acree2021; @lehmann2008book]. This is useful in cases where population parameters are unknown or unable to be reasonably measured. For example, some possible research questions that would require statistical methods include “What percentage of Minnesotan residents agree with the proposed legislation?,” or “Does the placement of higher margin products on the middle shelf of a store aisle lead to more profits than placement on top and bottom shelves?” These types of questions can be investigated using statistical inference, although the approach differs by the specific method of inference.

Implicitly, many statistical methods revolve around the concept of a hypothesis test [@neyman1933; @acree2021]. These tests are designed to determine whether or not there is evidence to support a claim. Data is evaluated against a null hypothesis, which is typically a baseline assumption or a specific condition. If the condition is deemed unlikely to occur by chance, then there is evidence to support its complement, called the alternative hypothesis. The strength of evidence can be quantified through summaries such as p-values and Bayes factors [@held2018], leading to decisions about what interpretation or decision to make regarding the results of the method.

@@

The use of visualizations is a valuable method for understanding the structure and patterns in data [@tukeyEDA], but they can also guide formal analyses [@loy2015; @loy2016; @loy2017]. Visual inference is where hypotheses can be tested by human perception instead of mathematical formulation. For example, Q-Q plots are a common diagnostic tool for checking normality assumptions [@loy2016]. While methods like the Shapiro-Wilk test [@shapiro1965] exist for testing normality, visual inference can be applied when assumptions are violated or tests do not exist.

@@

During the early development of Frequentist statistics, philosophical debates between R. A. Fisher and J. Neyman led to the distinction between "inductive inference" and "inductive behavior" [@lehmann2008]. Inductive inferences, as preferred by Fisher, involves logical rational to develop models. On the other hand, Neyman's inductive behavior addresses the additional component of randomization through hypothesis testing. Both arguments start with the assumption that the random variable in question is independent and identically distributed. However, the two statisticians differed on the interpretations and approaches, with Fisher favoring intuition and Neyman favoring proofs. The methodology was similar between them, emphasizing specific approaches and scope of inference.

@@

```{r}
#| label: fig-freq-eval-methods
#| fig-cap: "Frequentist evaluation methods using test statistics and confidence intervals under a null hypothesis that uses a standard normal distribution with a sample size of 100. Panel (a) shows the area under the sampling distribution for where $Z=\\frac{\\bar{X}}{1/\\sqrt{100}}$ would reject the null hypothesis. Panel (b) provides 95% confidence intervals for sample means across multiple samples."
#| fig-subcap: 
#| - "Probability of observing test statistic under null hypothesis."
#| - "95% confidence interval for multiple repeated samples."
#| out-width: 90%
#| fig-height: 4
#| layout-ncol: 2

set.seed(20244)
ggplot() + 
  stat_function(fun = \(x)(dnorm(x, sd = 1/sqrt(100))),
                xlim = c(-4/sqrt(100), 4/sqrt(100))) + 
  stat_function(fun = \(x)(dnorm(x, sd = 1/sqrt(100))),
                xlim = c(qnorm(0.975, sd = 1/sqrt(100)), 4/sqrt(100)),
                geom = 'area',
                fill = 'red', color = 'black') +
  stat_function(fun = \(x)(dnorm(x, sd = 1/sqrt(100))),
                xlim = c(-qnorm(0.975, sd = 1/sqrt(100)), -4/sqrt(100)),
                geom = 'area',
                fill = 'red', color = 'black') +
  labs(x = 'Z', y = expression(phi(Z))) + 
  scale_x_continuous(limits = c(-4/sqrt(100), 4/sqrt(100))) +
  theme(aspect.ratio = 1)

freq.ci <- data.frame()
for(i in 1:100){
  t.test.save <- t.test(rnorm(100))
  tmp <- data.frame(mean = t.test.save$estimate,
                    lcl = t.test.save$conf.int[1],
                    ucl = t.test.save$conf.int[2],
                    id = i)
  freq.ci <- bind_rows(freq.ci, tmp)
  rm(t.test.save)
  rm(tmp)
}
freq.ci %>% 
  mutate(contain0 = ifelse(0 > lcl & 0 <= ucl, 'yes', 'no')) %>% 
  ggplot(mapping = aes(x = mean, y = id, color = contain0)) + 
  geom_point() + 
  geom_errorbar(aes(xmin = lcl, xmax = ucl)) + 
  scale_color_brewer(palette = 'Set1') +
  labs(x = 'Z', y = 'Sample ID', color = 'CI contain zero?') + 
  geom_vline(xintercept = 0, color = 'red', linetype = 'dashed') + 
  theme(aspect.ratio = 1, legend.position = 'none')
```

@@

```{r}
#| label: fig-bayes-binom
#| fig-cap: "Posterior distributions for probability of success of a Binomial distriubtion using Beta and uniform prior distributions. Data was generated so that 13 out of 20 trials were successful with a true probability of 0.7, which is denoted by the red line. "
#| fig-height: 4
#| fig-align: center
#| out-width: 90%

@@

tibble(
  x = seq(0,1, l = 1000)
) %>% 
  mutate(fx.u = dbeta(x, 13+1, 20-13+1),
         fx.b = dbeta(x, 13+alpha+1, 20-13+beta+1),
         L = choose(20,13)*x^(13)*(1-x)^(20-13)) %>% 
  ggplot(mapping = aes(x = x)) + 
  geom_line(aes(y = fx.u, linetype = 'Uniform')) + 
  geom_line(aes(y = fx.b, linetype = 'Beta')) +
  geom_vline(xintercept = 0.7, color = 'red') +
  scale_linetype_manual(values = prior) + 
  labs(x = 'p', y = 'Posterior density', linetype = 'Prior') + 
  theme(aspect.ratio = 1/2)

@@

wild <- read_csv('wild23-25.csv', show_col_types = F) %>% 
  mutate(pt_diff = abs(gf-ga))
wild %>% 
  group_by(season, result, location) %>% 
  count() %>% 
  pivot_wider(names_from = location, values_from = n) %>% 
  mutate(season = paste0(season-1, '-', season)) %>% 
  knitr::kable(caption = 'Data from the 2023-24 and 2024-25 seasons for the Minnesota Wild. 0 represents losses and 1 represents wins.', align = 'c')
```

@@

```{r}
#| label: fig-true-props
#| fig-height: 4
#| fig-width: 6
#| fig-cap: "Observed proportion of wins for the Minnesota Wild."

wild %>% 
  group_by(season, location, result, pt_diff) %>% 
  count() %>% 
  pivot_wider(names_from = result, names_prefix = 'result',
              values_from = n, values_fill = 0) %>% 
  mutate(propwin = result1 / (result0+result1),
         season = map_chr(season, \(x)(paste0(season-1, '-', season)))) %>% 
  filter(!(result0 == 0 & result1 == 0)) %>% 
  ggplot(mapping = aes(x = pt_diff, y = propwin, color = location)) + 
  geom_line() + 
  geom_point() + 
  scale_y_continuous(limits = c(0,1)) + 
  labs(x = 'Point differential',
       y = 'Proportion of wins') +
  scale_color_manual(values = c('home' = '#154734', 'away' = '#A6192E')) +
  facet_wrap(~season) + 
  theme_bw() + 
  theme(aspect.ratio = 1) 
```

@@

Fitting a Frequentist generalized linear model (GLM) as shown in @eq-freq-glm reveals that the point differential is the only significant term (@tbl-anova), which means that the Wild do not seem to have a winning advantage by playing on home-ice. The only conclusion from the this analysis is that there is evidence that the Minnesota Wild are less likely to win games for larger point differentials than for smaller point differentials. In other words,

@@

```{r}
#| label: tbl-coef-compare

freq.coef <- summary(mod.freq)$coefficient[,1:2] %>% 
  as.data.frame() %>% 
  rownames_to_column() %>% 
  mutate(Frequentist = map2_chr(Estimate, `Std. Error`, \(x,y)(glue::glue('{round(x,2)} ({round(y, 2)})'))))

bayes.u.coef <- fixef(bayes.unif)[,1:2] %>% 
  as.data.frame() %>% 
  rownames_to_column() %>% 
  mutate(Bayes.unif = map2_chr(Estimate, Est.Error, \(x,y)(glue::glue('{round(x,2)} ({round(y, 2)})'))))

bayes.n.coef <- fixef(bayes.priors)[,1:2] %>% 
  as.data.frame() %>% 
  rownames_to_column() %>% 
  mutate(Bayes.normal = map2_chr(Estimate, Est.Error, \(x,y)(glue::glue('{round(x,2)} ({round(y, 2)})'))))

data.frame(
  Term = freq.coef$rowname,
  Freq = freq.coef$Frequentist,
  `Uniform priors` = bayes.u.coef$Bayes.unif,
  `Informative priors` = bayes.n.coef$Bayes.normal
) %>% 
  # rownames_to_column(var = 'Term') %>% 
  knitr::kable(digits = 3, caption = 'Coefficients and standard errors of Frequentist and Bayesian methods for GLMs fit to the Minnesota Wild data.', col.names = c('Term', 'Frequentist GLM', 'Bayes GLM (Uniform Priors)', 'Bayes GLM (Normal Priors)'), align = 'c')
